Supplement to Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida

### 2021-11-04 supplement.Rmd
NW comments added

---
title: "Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay. Port operations were primarily for export of phosphate production by the plant.  Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].  Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.  In 1997, 204 million liters of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes [@DiPinto01].

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to the Florida Department of Environmental Protection (FDEP).  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.  For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers. comm. D. Eckenrod to USEPA, Nov. 28, 2001).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. In other event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication. Around the same time, 939 million liters of wastewater from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

@@

For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.

@@

Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.  Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17]. Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).  This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant. Seagrasses and macroalgae were identified and Braun-Blanquet abundances were estimated within a 0.25 m$^2$ quadrat at 10m distances along each transect.  Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.  Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.  Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria).  Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.

@@

```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics(here('figs/wqtrnds-supp.jpeg'))
```

```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics('figs/wqtrndsadd-supp.jpeg')
```

>>>>

---
title: "Supplement to Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C. Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W. Chin (diwchin@gmail.com), Emma E. Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J. Phlips (phlips@ufl.edu), Gary E. Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T. Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

@@

The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay. Port operations were primarily for export of phosphate production by the plant.  Numerous environmental issues were observed in these early years, including suspected surface water contamination from Piney Point in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].  Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.  In 1997, 204 million liters of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes [@DiPinto01].

The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to the Florida Department of Environmental Protection (FDEP).  Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of water on-site to minimize environmental impacts.  Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.  For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of discharge from Piney Point to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers. comm. D. Eckenrod to USEPA, Nov. 28, 2001).  Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11]. In other event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill. @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication. Around the same time, 939 million liters of water from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].  Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.

@@

For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.  This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.  Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).  Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.  As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.

@@

Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.  Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17]. Because of the time-sensitive nature of the potential impacts of discharge from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).  This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant. Seagrasses and macroalgae were identified and Braun-Blanquet abundances were estimated within a 0.25 m$^2$ quadrat at 10m distances along each transect.  Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.  Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.  Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria).  Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.

@@

```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics(here('figs/wqtrnds-supp.jpeg'))
```

```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).  Observations are aggregated by week and within assessment areas shown in Figure 1a.  Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas. Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
knitr::include_graphics('figs/wqtrndsadd-supp.jpeg')
```
